Stochastic Variance-Reduced Policy Gradient

### 2018-02-05 svrpg.tex
updated definition of variance and notation

The idea of SVRG is to alternate full and stochastic gradient updates. 
Each $m = \mathcal{O}(N)$ iterations, a snapshot $\widetilde{\vtheta}$ of the current parameter is saved together with its full gradient $\nabla f(\widetilde{\vtheta}) = \frac{1}{N} \sum_i z(x_i|\widetilde{\vtheta})$.
Between snapshots, the parameter is updated with a gradient estimate corrected using stochastic gradient, for any $t \in \{0,\ldots,m-1\}$
\begin{equation}\label{E:svrg.gradient.correction}
        \blacktriangledown f(\vtheta_{t+1}) = \nabla f(\tilde{\vtheta}) + \nabla z(x_i | \vtheta_t) - \nabla z(x_i | \tilde{\vtheta}),
\end{equation} 
where $x_i$ is sampled uniformly at random from $\mathcal{D}_N$ (\ie $x_i \sim \mathcal{U}(\mathcal{D}_N)$). The corrected gradient $\blacktriangledown f(\vtheta)$ is an unbiased estimate of $\nabla f(\vtheta)$, and it is able to control the variance introduced by sampling even with a fixed step size, without resorting to plain full gradient. The resulting IFO complexity is $\mathcal{O}(N+\nicefrac{\sqrt{N}}{\epsilon})$ \cite{reddi2016stochastic}.

@@

We can now introduce Stochastic Variance-Reduced Policy Gradient (SVRPG).
% For simplicity, we focus on the REINFORCE estimator but the extension to G(PO)MDP is straightforward.
Between snapshots, the SVRPG estimate is corrected using importance weighting, for any $t \in \{0,\ldots,m-1\}$
\begin{align*}
        \blacktriangledown J(\vtheta_{t+1}) &= \wh{\nabla}_N J(\wt{\vtheta}) + g(\tau_i|\vtheta_t) - \omega(\tau_i|\vtheta_t, \wt{\vtheta}) g(\tau_i|\wt{\vtheta})
\end{align*}
% \begin{align*}
% 	\blacktriangledown J(\vtheta) &= \frac{1}{N}\sum_{j=0}^{N-1}\nabla\log\pi(\tau_j \vert \tilde{\vtheta})\Reward(\tau_j) \\
% 		&+ \nabla\log\pi(\tau_i \vert \vtheta)\Reward(\tau_i) 
% 		- \omega(\tau_i)\nabla\log\pi(\tau_i \vert \tilde{\vtheta})\Reward(\tau_i),
% \end{align*}
where $\wh{\nabla}_N J(\wt{\vtheta})$ is as in~\eqref{E:policygradient.estimate} where $\mathcal{D}_N$ is sampled using the snapshot policy $\pi_{\wt{\vtheta}}$, $\tau_i$ is sampled from the current policy $\pi_{\vtheta_t}$, and $\omega(\tau|\vtheta_t, \wt{\vtheta}) = \frac{\pi(\tau|\wt{\vtheta})}{\pi(\tau|\vtheta_t)}$ is an importance weight from $\pi_{\vtheta_t}$ to the snapshot policy $\pi_{\wt{\vtheta}}$. 
Our update is still fundamentally on-policy, since the weighting concerns only the correction term. However, this partial ``off-policyness'' is enough to introduce variance. This is a well-known issue of importance sampling~\citep[\eg][]{thomas2015high}. To mitigate it, we use mini-batches of trajectories of size $B \ll N$ to average the correction, \ie
\begin{align}\label{E:svrpg.estimate.batch}
        \blacktriangledown J(\vtheta_{t+1}) &= \wh{\nabla}_N J(\wt{\vtheta})\\ \notag
                                            & \quad{} + \frac{1}{B} \sum_{i=0}^{B-1} \left[
        g(\tau_i|\vtheta_t) - \omega(\tau_i|\vtheta_t, \wt{\vtheta}) g(\tau_i|\wt{\vtheta})
        \right].
\end{align}
%\begin{align*}
%&\blacktriangledown J(\vtheta) = \frac{1}{N}\sum_{j=0}^{N-1}\nabla\log\pi(\tau_j \vert \tilde{\vtheta})\Reward(\tau_j) +\\
%&\quad\frac{1}{B}\left[\sum_{i=0}^{B-1}
%\nabla\log\pi(\tau_i \vert \vtheta)\Reward(\tau_i) 
%- \omega(\tau_i)\nabla\log\pi(\tau_i \vert \tilde{\vtheta})\Reward(\tau_i)\right].
%\end{align*}
The use of mini-batches is a common practice in SVRG since it can yield a performance improvement even in the supervised case~\citep{harikandeh2015stopwasting,konevcny2016mini}. It is easy to show that the SVRPG estimator has the following, desirable properties (see App.~\ref{app:proofs}):
\begin{lemma}
Denote by $\wh{\nabla}_N^{\textsc{RF}} J(\vtheta)$ the REINFORCE estimator in~\eqref{E:policygradient.estimate} and let $\vtheta^* \in \argmin_{\vtheta} \{J(\vtheta)\}$. Then, the SVRG estimate in~\eqref{E:svrpg.estimate.batch} is \emph{unbiased}
\[
\mathop{\mathbb{E}}
\left[\blacktriangledown J(\vtheta)\right] = \gradJ{\vtheta}.
\]
and regardless of the mini-batch size $B$:
\[
	\mathbb{V}ar\left[\gradBlack{\vtheta^*}\right] = 
    \mathbb{V}ar\left[\wh{\nabla}_N^{\textsc{RF}}J(\vtheta^*)\right].
\]
\todopir{Definition of Var??}
\end{lemma}

@@

\textit{2) FG Approximation.}
Since we cannot compute an exact full gradient, we require the variance of the estimator to be bounded.
This assumption is in spirit similar to the one in~\citep{harikandeh2015stopwasting}.
	\begin{restatable}[On the variance of the gradient estimator]{assumption}{varreinforce}\label{ass:REINFORCE}
		There is a constant $V<\infty$ such that, for any policy $\pol$:
		\[
		\Tr\mathbb{V}ar\left(\gradApp{\vtheta}{1}\right) \leq \VARRF.
		\]
        \todopir{Why 1 as subscript?}
	\end{restatable}

\textit{3) Non-stationarity.} 
Similarly to what done in SL~\citep{cortes2010learning}, we require the variance of the importance weight to be bounded.
	\begin{restatable}[On the variance of importance weights]{assumption}{varweights}\label{ass:M2}
		There is a constant $W<\infty$ such that, for each pair of policies encountered in Alg.~\ref{alg:svrpg} and for each possible trajectory,
		\[
                \mathbb{V}ar\left(\omega(\tau| \vtheta_1, \vtheta_2)\right) \leq \VARIS, \quad \forall \vtheta_1,\vtheta_2 \in \realspace^d , \tau \sim \pi_{\vtheta_1}.
		\]
	\end{restatable}
Differently from Ass.~\ref{ass:bounded_score} and ~\ref{ass:REINFORCE}, Ass.~\ref{ass:M2} can be enforced by a proper handling of the epoch size $m$.

@@

\subsection*{Definitions}
We give some additional definitions which will be useful in the proofs.
\begin{definition}
For a random variable $X$:
\begin{align*}
	&\mathbb{E}_{t\vert s}\left[X\right] \coloneqq 
		\mathop{\mathbb{E}}_{\substack{\tau_j\sim\pi(\cdot\vert\tilde{\vtheta}^s)\forall j \in {\it {\bf N}} \\ \tau_{i,h}\sim\pi(\cdot\vert\vtheta^{s+1}_i) \forall h \in {\it {\bf B}}, \text{ for $i=0,\dots,t-1$}}}{\left[X \vert \tilde{\vtheta^s}\right]} \\
	&\coloneqq \EVV[\tau_j\sim\pi(\cdot\vert\tilde{\vtheta}^s)\forall j \in {\it {\bf N}}]{
			\EVV[\tau_{0,h}\sim\pi(\cdot\vert\vtheta_0^{s+1}) \forall h \in {\it {\bf B}}]
				{\dots
					\EVV[\tau_{t,h}\sim\pi(\cdot\vert\vtheta_t^{s+1})\forall h \in {\it {\bf B}}]
						{X\vert\vtheta_t^{s+1}}
				 \dots
			\vert\vtheta_0^{s+1}}
		\vert\tilde{\vtheta}^s},
\end{align*}
where the sequence $\tilde{\vtheta}^s,\vtheta_0^{s+1},\dots,\vtheta_t^{s+1}$ is defined in Algorithm \ref{alg:svrpg}, ${\it {\bf N}} = [0,\dots,N)$, and ${\it {\bf B}} = [0,\dots,B)$. 
\end{definition}
Intuitively, the $\Ets{\cdot}$ operator computes the expected value with respect to the sampling of trajectories from the snapshot $\tilde{\vtheta}^s$ up to the $t$-th iteration included. Note that the order in which expected values are taken is important since each $\vtheta_{t}^{s+1}$ is function of previously sampled trajectories and is used to sample new ones.

@@

% \zerovar*
\begin{restatable}{lemma}{zerovar}\label{lemma:zerovar}
At convergence, the variance of the SVRPG estimator is equal to the variance of the REINFORCE estimator, regardless of the mini-batch size $B$:
\[
	\mathbb{V}ar\left[\gradBlack{\vtheta^*}\right] = 
	\mathbb{V}ar\left[\gradApp{\vtheta^*}{N}\right].
\]
\end{restatable}
\begin{proof}
As $\vtheta\to\vtheta^*$, also $\tilde{\vtheta}\to\vtheta^*$. Hence, by continuity of $J(\vtheta)$:
\begin{align*}
\Var\left[\gradBlack{\vtheta}\right] &\to \Var\left[\gradApp{\vtheta^*}{N}\right] + \frac{1}{B}\Var\left[\score{\vtheta^*}{\tau}R(\tau) - \cancel{\omega(\tau)}\score{\vtheta^*}{\tau}R(\tau)\right] \\
&= \Var\left[\gradApp{\vtheta^*}{N}\right].
\end{align*}
Note that it is important that the trajectories used in the second and the third term are the same for the variance to vanish.
\end{proof}

@@

+\frac{1}{N}\Tr\Varts[]{\gradApp{\tilde{\vtheta}^s}{1}}

@@

	+\frac{1}{N}\Tr\Varts[]{\gradApp{\tilde{\vtheta}^s}{1}} \nonumber\\

@@

%Lemma 0
\begin{restatable}[]{lemma}{auxzero}\label{lemma:aux0}
Under Assumption \ref{ass:bounded_score}, for any function $f(\vtheta_t^{s+1})$ which is deterministic for a fixed $\vtheta_t^{s+1}$:
\begin{align*}
\left|\Ets[t]{\dotprod{\gradBlack{\vtheta_t^{s+1}}}{f(\vtheta_t^{s+1})}}
-\Ets{\dotprod{\gradJ{\vtheta_t^{s+1}}}{f(\vtheta_t^{s+1})}}
\right|
\leq
\frac{1}{2N}\Tr\Varts[]{\gradApp{\tilde{\vtheta}^s}{1}} +\frac{1}{2}\Ets[t-1]{\norm[]{f(\vtheta_t^{s+1})}^2}
\end{align*}
\end{restatable}
\begin{proof}
\begin{align}
	\Ets{\dotprod{\gradBlack{\vtheta_t^{s+1}}}{f(\vtheta_t^{s+1})}}
	&=
	\Ets{\dotprod{\gradIdeal{\vtheta_t^{s+1}}}{f(\vtheta_t^{s+1})}} +
	\Ets[t-1]{\dotprod{e_s}{f(\vtheta_t^{s+1})}} \label{eq:6}\\
	&=
	\Ets{\dotprod{\gradJ{\vtheta_t^{s+1}}}{f(\vtheta_t^{s+1})}} +
	\Ets[t-1]{\dotprod{e_s}{f(\vtheta_t^{s+1})}} \label{eq:7}\\
	&=
	\Ets{\dotprod{\gradJ{\vtheta_t^{s+1}}}{f(\vtheta_t^{s+1})}} \nonumber\\
	&\qquad+
	\dotprod{\Ets{e_s}}{\Ets{f(\vtheta_t^{s+1})}}
	+\Tr\Covts[t-1]{\gradApp{\tilde{\vtheta}^s}{N}}{f(\vtheta_t^{s+1})}  \nonumber\\
	&= 
	\Ets{\dotprod{\gradJ{\vtheta_t^{s+1}}}{f(\vtheta_t^{s+1})}} \nonumber\\
	&\qquad+
	\Tr\Covts[t-1]{\gradApp{\tilde{\vtheta}^s}{N}}{f(\vtheta_t^{s+1})} \label{eq:8}
\end{align}
Hence:
\begin{align}
	\left|\Ets{\dotprod{\gradBlack{\vtheta_t^{s+1}}}{f(\vtheta_t^{s+1})}}
	\right.&-\left.\Ets{\dotprod{\gradJ{\vtheta_t^{s+1}}}{f(\vtheta_t^{s+1})}}\right| 
	=
	\left|\Tr\Covts[t-1]{\gradApp{\tilde{\vtheta}^s}{N}}{f(\vtheta_t^{s+1})}\right|  
	\nonumber\\
	&\leq
	\Tr\left|\Covts[t-1]{\gradApp{\tilde{\vtheta}^s}{N}}{f(\vtheta_t^{s+1})}\right| \nonumber \\
	%
	&\leq
	\Tr\left(\sqrt{\Varts[]{\gradApp{\tilde{\vtheta}^s}{N}}}\circ\sqrt{\Varts[t-1]{f(\vtheta_t^{s+1})}}\right) \nonumber\\
	%
	&\leq	
	\frac{\alpha_t}{2}\Tr\Varts[]{\gradApp{\tilde{\vtheta}^s}{N}} +\frac{\alpha_t}{2}\Tr\Varts[t-1]{f(\vtheta_t^{s+1})}\label{eq:9}\\
	%
	&=
	\frac{1}{2N}\Tr\Varts[]{\gradApp{\tilde{\vtheta}^s}{1}} +\frac{1}{2}\Tr\Varts[t-1]{f(\vtheta_t^{s+1})} \label{eq:10}\\
	%
	&\leq
	\frac{1}{2N}\Tr\Varts[]{\gradApp{\tilde{\vtheta}^s}{1}} +\frac{1}{2}\Ets[t-1]{\norm[]{f(\vtheta_t^{s+1})}^2},
	\nonumber
\end{align}
where (\ref{eq:6}) is from Definition \ref{def:ideal}; (\ref{eq:7}) is from the fact that $\gradIdeal{\vtheta_t^{s+1}}$ is both unbiased and independent from $f(\vtheta_t^{s+1})$ \wrt the sampling at time $t$ alone, which is not true for $\gradBlack{\vtheta_t^{s+1}}$; (\ref{eq:8}) is from $\Ets{e_s}=0$; (\ref{eq:9}) is from Young's inequality; and (\ref{eq:10}) is from the definition of $\gradApp{\vtheta}{N}$.
\end{proof}

>>>>

The idea of SVRG is to alternate full and stochastic gradient updates. 
Each $m = \mathcal{O}(N)$ iterations, a snapshot $\widetilde{\vtheta}$ of the current parameter is saved together with its full gradient $\nabla f(\widetilde{\vtheta}) = \frac{1}{N} \sum_i z(x_i|\widetilde{\vtheta})$.
Between snapshots, the parameter is updated with a gradient estimate corrected using stochastic gradient, for any $t \in \{0,\ldots,m-1\}$
\begin{equation}\label{E:svrg.gradient.correction}
        \blacktriangledown f(\vtheta_{t}) = \nabla f(\tilde{\vtheta}) + \nabla z(x_i | \vtheta_t) - \nabla z(x_i | \tilde{\vtheta}),
\end{equation} 
where $x_i$ is sampled uniformly at random from $\mathcal{D}_N$ (\ie $x_i \sim \mathcal{U}(\mathcal{D}_N)$).
Note that $t=0$ corresponds to a FG step (\ie $\blacktriangledown f(\vtheta_0) = \nabla f(\wt{\vtheta})$) since $\vtheta_0 := \wt{\vtheta}$.
The corrected gradient $\blacktriangledown f(\vtheta)$ is an unbiased estimate of $\nabla f(\vtheta)$, and it is able to control the variance introduced by sampling even with a fixed step size, without resorting to plain full gradient. The resulting IFO complexity is $\mathcal{O}(N+\nicefrac{\sqrt{N}}{\epsilon})$ \cite{reddi2016stochastic}.

@@

We can now introduce Stochastic Variance-Reduced Policy Gradient (SVRPG).
% For simplicity, we focus on the REINFORCE estimator but the extension to G(PO)MDP is straightforward.
Between snapshots, the SVRPG estimate is corrected using importance weighting, for any $t \in \{0,\ldots,m-1\}$
\begin{align*}
        \blacktriangledown J(\vtheta_{t}) &= \wh{\nabla}_N J(\wt{\vtheta}) + g(\tau_i|\vtheta_t) - \omega(\tau_i|\vtheta_t, \wt{\vtheta}) g(\tau_i|\wt{\vtheta})
\end{align*}
% \begin{align*}
% 	\blacktriangledown J(\vtheta) &= \frac{1}{N}\sum_{j=0}^{N-1}\nabla\log\pi(\tau_j \vert \tilde{\vtheta})\Reward(\tau_j) \\
% 		&+ \nabla\log\pi(\tau_i \vert \vtheta)\Reward(\tau_i) 
% 		- \omega(\tau_i)\nabla\log\pi(\tau_i \vert \tilde{\vtheta})\Reward(\tau_i),
% \end{align*}
where $\wh{\nabla}_N J(\wt{\vtheta})$ is as in~\eqref{E:policygradient.estimate} where $\mathcal{D}_N$ is sampled using the snapshot policy $\pi_{\wt{\vtheta}}$, $\tau_i$ is sampled from the current policy $\pi_{\vtheta_t}$, and $\omega(\tau|\vtheta_t, \wt{\vtheta}) = \frac{\pi(\tau|\wt{\vtheta})}{\pi(\tau|\vtheta_t)}$ is an importance weight from $\pi_{\vtheta_t}$ to the snapshot policy $\pi_{\wt{\vtheta}}$. 
Similarly to SVRG, we have that $\vtheta_0 := \wt{\vtheta}$ and the update is a FG step.
Our update is still fundamentally on-policy, since the weighting concerns only the correction term. However, this partial ``off-policyness'' is enough to introduce variance. This is a well-known issue of importance sampling~\citep[\eg][]{thomas2015high}. To mitigate it, we use mini-batches of trajectories of size $B \ll N$ to average the correction, \ie
\begin{align}\label{E:svrpg.estimate.batch}
        \blacktriangledown J(\vtheta_{t}) &:= v_t= \wh{\nabla}_N J(\wt{\vtheta})\\ \notag
                                            & \quad{} + \frac{1}{B} \sum_{i=0}^{B-1} \left[
        g(\tau_i|\vtheta_t) - \omega(\tau_i|\vtheta_t, \wt{\vtheta}) g(\tau_i|\wt{\vtheta})
        \right].
\end{align}
%\begin{align*}
%&\blacktriangledown J(\vtheta) = \frac{1}{N}\sum_{j=0}^{N-1}\nabla\log\pi(\tau_j \vert \tilde{\vtheta})\Reward(\tau_j) +\\
%&\quad\frac{1}{B}\left[\sum_{i=0}^{B-1}
%\nabla\log\pi(\tau_i \vert \vtheta)\Reward(\tau_i) 
%- \omega(\tau_i)\nabla\log\pi(\tau_i \vert \tilde{\vtheta})\Reward(\tau_i)\right].
%\end{align*}
The use of mini-batches is a common practice in SVRG since it can yield a performance improvement even in the supervised case~\citep{harikandeh2015stopwasting,konevcny2016mini}. It is easy to show that the SVRPG estimator has the following, desirable properties (see App.~\ref{app:proofs}):
\begin{lemma}
Denote by $\wh{\nabla}_N^{\textsc{RF}} J(\vtheta)$ the REINFORCE estimator in~\eqref{E:policygradient.estimate} and let $\vtheta^* \in \argmin_{\vtheta} \{J(\vtheta)\}$. Then, the SVRG estimate in~\eqref{E:svrpg.estimate.batch} is \emph{unbiased}
\[
\mathop{\mathbb{E}}
\left[\blacktriangledown J(\vtheta)\right] = \gradJ{\vtheta}.
\]
and regardless of the mini-batch size $B$:
\[
	\Var\left[\gradBlack{\vtheta^*}\right] = 
    \Var\left[\wh{\nabla}_N^{\textsc{RF}}J(\vtheta^*)\right].
\]
\end{lemma}
For any vector $\mathbf{x}$, we use $\Var[\mathbf{x}]$ to denote the trace of the covariance matrix, \ie $\Tr\EVV[]{(\mathbf{x}-\EVV[]{\mathbf{x}})(\mathbf{x}-\EVV[]{\mathbf{x}})^T}$.

@@

\textit{2) FG Approximation.}
Since we cannot compute an exact full gradient, we require the variance of the estimator to be bounded.
This assumption is in spirit similar to the one in~\citep{harikandeh2015stopwasting}.
	\begin{restatable}[On the variance of the gradient estimator]{assumption}{varreinforce}\label{ass:REINFORCE}
		There is a constant $V<\infty$ such that, for any policy $\pol$:
		\[
			\Var\left[g(\cdot\vert\vtheta)\right] \leq \VARRF.
		\]
	\end{restatable}

\textit{3) Non-stationarity.} 
Similarly to what done in SL~\citep{cortes2010learning}, we require the variance of the importance weight to be bounded.
	\begin{restatable}[On the variance of importance weights]{assumption}{varweights}\label{ass:M2}
		There is a constant $W<\infty$ such that, for each pair of policies encountered in Alg.~\ref{alg:svrpg} and for each possible trajectory,
		\[
                \mathbb{V}ar\left[\omega(\tau| \vtheta_1, \vtheta_2)\right] \leq \VARIS, \quad \forall \vtheta_1,\vtheta_2 \in \realspace^d , \tau \sim \pi_{\vtheta_1}.
		\]
	\end{restatable}
Differently from Ass.~\ref{ass:bounded_score} and ~\ref{ass:REINFORCE}, Ass.~\ref{ass:M2} can be enforced by a proper handling of the epoch size $m$.

@@

\subsection*{Definitions}
We give some additional definitions which will be useful in the proofs.

\begin{definition}
For a random variable $X$:
\begin{align*}
	&\mathbb{E}_{t\vert s}\left[X\right] \coloneqq 
		\mathop{\mathbb{E}}_{\substack{\tau_j\sim\pi(\cdot\vert\tilde{\vtheta}^s)\forall j \in {\it {\bf N}} \\ \tau_{i,h}\sim\pi(\cdot\vert\vtheta^{s+1}_i) \forall h \in {\it {\bf B}}, \text{ for $i=0,\dots,t-1$}}}{\left[X \vert \tilde{\vtheta^s}\right]} \\
	&\coloneqq \EVV[\tau_j\sim\pi(\cdot\vert\tilde{\vtheta}^s)\forall j \in {\it {\bf N}}]{
			\EVV[\tau_{0,h}\sim\pi(\cdot\vert\vtheta_0^{s+1}) \forall h \in {\it {\bf B}}]
				{\dots
					\EVV[\tau_{t,h}\sim\pi(\cdot\vert\vtheta_t^{s+1})\forall h \in {\it {\bf B}}]
						{X\vert\vtheta_t^{s+1}}
				 \dots
			\vert\vtheta_0^{s+1}}
		\vert\tilde{\vtheta}^s},
\end{align*}
where the sequence $\tilde{\vtheta}^s,\vtheta_0^{s+1},\dots,\vtheta_t^{s+1}$ is defined in Algorithm \ref{alg:svrpg}, ${\it {\bf N}} = [0,\dots,N)$, and ${\it {\bf B}} = [0,\dots,B)$. 
\end{definition}
Intuitively, the $\Ets{\cdot}$ operator computes the expected value with respect to the sampling of trajectories from the snapshot $\tilde{\vtheta}^s$ up to the $t$-th iteration included. Note that the order in which expected values are taken is important since each $\vtheta_{t}^{s+1}$ is function of previously sampled trajectories and is used to sample new ones.

\begin{definition}
For random vectors X, Y:
\begin{align*}
	\Covts{X}{Y} &\coloneqq \Tr\left(\Ets{(X-\Ets{X})(Y-\Ets{Y})^T}\right), \\
	\Varts{X} &\coloneqq \Covts{X}{Y},
\end{align*}
where $\Tr(\cdot)$ denotes the trace of a matrix.
\end{definition}

@@

% \zerovar*
\begin{restatable}{lemma}{zerovar}\label{lemma:zerovar}
At convergence, the variance of the SVRPG estimator is equal to the variance of the REINFORCE estimator, regardless of the mini-batch size $B$:
\[
	\Var\left[\gradBlack{\vtheta^*}\right] = 
	\Var\left[\gradApp{\vtheta^*}{N}\right].
\]
\end{restatable}
\begin{proof}
As $\vtheta\to\vtheta^*$, also $\tilde{\vtheta}\to\vtheta^*$. Hence, by continuity of $J(\vtheta)$:
\begin{align*}
\Var\left[\gradBlack{\vtheta}\right] &\to \Var\left[\gradApp{\vtheta^*}{N}\right] + \frac{1}{B}\Var\left[\score{\vtheta^*}{\tau}R(\tau) - \cancel{\omega(\tau)}\score{\vtheta^*}{\tau}R(\tau)\right] \\
&= \Var\left[\gradApp{\vtheta^*}{N}\right].
\end{align*}
Note that it is important that the trajectories used in the second and the third term are the same for the variance to vanish.
\end{proof}

@@

+\frac{1}{N}\Varts[]{\gradApp{\tilde{\vtheta}^s}{1}}

@@

	+\frac{1}{N}\Varts[]{\gradApp{\tilde{\vtheta}^s}{1}} \nonumber\\

@@

%Lemma 0
\begin{restatable}[]{lemma}{auxzero}\label{lemma:aux0}
Under Assumption \ref{ass:bounded_score}, for any function $f(\vtheta_t^{s+1})$ which is deterministic for a fixed $\vtheta_t^{s+1}$:
\begin{align*}
\left|\Ets[t]{\dotprod{\gradBlack{\vtheta_t^{s+1}}}{f(\vtheta_t^{s+1})}}
-\Ets{\dotprod{\gradJ{\vtheta_t^{s+1}}}{f(\vtheta_t^{s+1})}}
\right|
\leq
\frac{1}{2N}\Varts[]{\gradApp{\tilde{\vtheta}^s}{1}} +\frac{1}{2}\Ets[t-1]{\norm[]{f(\vtheta_t^{s+1})}^2}
\end{align*}
\end{restatable}
\begin{proof}
\begin{align}
	\Ets{\dotprod{\gradBlack{\vtheta_t^{s+1}}}{f(\vtheta_t^{s+1})}}
	&=
	\Ets{\dotprod{\gradIdeal{\vtheta_t^{s+1}}}{f(\vtheta_t^{s+1})}} +
	\Ets[t-1]{\dotprod{e_s}{f(\vtheta_t^{s+1})}} \label{eq:6}\\
	&=
	\Ets{\dotprod{\gradJ{\vtheta_t^{s+1}}}{f(\vtheta_t^{s+1})}} +
	\Ets[t-1]{\dotprod{e_s}{f(\vtheta_t^{s+1})}} \label{eq:7}\\
	&=
	\Ets{\dotprod{\gradJ{\vtheta_t^{s+1}}}{f(\vtheta_t^{s+1})}} \nonumber\\
	&\qquad+
	\dotprod{\Ets{e_s}}{\Ets{f(\vtheta_t^{s+1})}}
	+\Covts[t-1]{\gradApp{\tilde{\vtheta}^s}{N}}{f(\vtheta_t^{s+1})}  \nonumber\\
	&= 
	\Ets{\dotprod{\gradJ{\vtheta_t^{s+1}}}{f(\vtheta_t^{s+1})}} \nonumber\\
	&\qquad+
	\Covts[t-1]{\gradApp{\tilde{\vtheta}^s}{N}}{f(\vtheta_t^{s+1})} \label{eq:8}
\end{align}
Hence:
\begin{align}
	\left|\Ets{\dotprod{\gradBlack{\vtheta_t^{s+1}}}{f(\vtheta_t^{s+1})}}
	\right.&-\left.\Ets{\dotprod{\gradJ{\vtheta_t^{s+1}}}{f(\vtheta_t^{s+1})}}\right| 
	=
	\left|\Covts[t-1]{\gradApp{\tilde{\vtheta}^s}{N}}{f(\vtheta_t^{s+1})}\right|  
	\nonumber\\
	&\leq
	\sqrt{\Varts[]{\gradApp{\tilde{\vtheta}^s}{N}}}\circ\sqrt{\Varts[t-1]{f(\vtheta_t^{s+1})}} \nonumber\\
	%
	&\leq	
	\frac{\alpha_t}{2}\Varts[]{\gradApp{\tilde{\vtheta}^s}{N}} +\frac{\alpha_t}{2}\Varts[t-1]{f(\vtheta_t^{s+1})}\label{eq:9}\\
	%
	&=
	\frac{1}{2N}\Varts[]{\gradApp{\tilde{\vtheta}^s}{1}} +\frac{1}{2}\Varts[t-1]{f(\vtheta_t^{s+1})} \label{eq:10}\\
	%
	&\leq
	\frac{1}{2N}\Varts[]{\gradApp{\tilde{\vtheta}^s}{1}} +\frac{1}{2}\Ets[t-1]{\norm[]{f(\vtheta_t^{s+1})}^2},
	\nonumber
\end{align}
where (\ref{eq:6}) is from Definition \ref{def:ideal}; (\ref{eq:7}) is from the fact that $\gradIdeal{\vtheta_t^{s+1}}$ is both unbiased and independent from $f(\vtheta_t^{s+1})$ \wrt the sampling at time $t$ alone, which is not true for $\gradBlack{\vtheta_t^{s+1}}$; (\ref{eq:8}) is from $\Ets{e_s}=0$; (\ref{eq:9}) is from Young's inequality; and (\ref{eq:10}) is from the definition of $\gradApp{\vtheta}{N}$.
\end{proof}

%Lemma 1
\begin{restatable}[]{lemma}{auxone}\label{lemma:aux1}
Under Assumptions \ref{ass:bounded_score} ans \ref{ass:M2}, the expected squared norm of the true gradient $\gradJ{\vtheta_t^{s+1}}$, for appropriate choices of $\alpha_t\geq0$ and $\beta_t>0$, can be bounded as follows:
\[
	\Ets{\norm[]{\gradJ{\vtheta_t^{s+1}}}^2} \leq
	\frac{R_{t+1}^{s+1} - R_t^{s+1}}{\Psi_t} + \frac{d_tV}{N\Psi_t}
	+\frac{f_tW}{b\Psi_t},
\]
	where
\begin{align*}
	&R_t^{s+1}\coloneqq \Ets{J(\vtheta_t^{s+1}) - c_t\norm[]{\vtheta_t^{s+1}-\tilde{\vtheta}^s}^2}, \\
	&c_{m-1} = 0, \\
	&c_t = c_{t+1}\left(\frac{3}{2}+\alpha_t\beta_t+\alpha_t^2\right) + \alpha_t\frac{L^2}{b}+\frac{\alpha_t^2L^3}{2}, \\
	&\Psi_t = \alpha_t\left(1-\frac{c_{t+1}}{\beta_t}-\frac{\alpha_tL}{2}-\alpha_tc_{t+1}\right), \\
	&d_t = \frac{\alpha_t}{2}\left(1+2c_{t+1}+\alpha_t^2L+2\alpha_t^2c_{t+1}\right), \\
	&f_t = \alpha_t^2\frac{G^2R^2(L+2c_{t+1})}{2(1-\gamma)^2}.
\end{align*}
In particular, the following constraints on $\alpha_t$ and $\beta_t$ are sufficient:
\begin{align*}
&0\leq\alpha_t < \frac{2(1-\nicefrac{c_{t+1}}{\beta_t})}{L+2c_{t+1}} \\
&\beta_t > c_{t+1}.
\end{align*}
\end{restatable}
\begin{proof}
	We have:
	\begin{align}
	\Ets{J(\vtheta_{t+1}^{s+1})} 
	&\geq \Ets{J(\vtheta_t^{s+1})+\dotprod{\gradJ{\vtheta_t^{s+1}}}{\vtheta_{t+1}^{s+1}-\vtheta_t^{s+1}} - \frac{L}{2}\norm[]{\vtheta_{t+1}^{s+1}-\vtheta_t^{s+1}}^2} \label{eq:11}\\
	&= \Ets{J(\vtheta_t^{s+1})+\alpha_t\dotprod{\gradJ{\vtheta_t^{s+1}}}{\gradBlack{\vtheta_t^{s+1}}} - \frac{\alpha_t^2L}{2}\norm[]{\gradBlack{\vtheta_t^{s+1}}}^2} \label{eq:12}\\
	&\geq
	\Ets{J(\vtheta_t^{s+1})+\alpha_t\norm[]{\gradJ{\vtheta_t^{s+1}}}^2 - \frac{\alpha_t^2L}{2}\norm[]{\gradBlack{\vtheta_t^{s+1}}}^2} \nonumber\\
	&\qquad-
	\frac{\alpha_t}{2N}\Varts[]{\gradApp{\tilde{\vtheta}^s}{1}} -\frac{\alpha_tL^2}{2}\Ets[t-1]{\norm[]{\vtheta_t^{s+1} - \tilde{\vtheta}^s}^2}, \label{eq:13}
\end{align}
where (\ref{eq:11}) is from the L-smoothness of $J(\vtheta)$, (\ref{eq:12}) is from the SVRPG update, and (\ref{eq:13}) is from Lemma \ref{lemma:aux0}.
